Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-15 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
CF changes incorp Ch 2 through 5

You have now developed your first hypothesis about the cause of the fire.  A radiant style, electric heater was turned on.  It was too close to the furniture and ignited the fabric on the sofa or chair.  You collect the remains of the heater for further examination to have a lab determine if the heater switch was in the ``on'' position.  Now you need to start testing your hypothesis.  Is everything consistent with the other fire damage?

@@

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover, or early growth stage, fires.  A fixed fire source footprint, a 30.5~cm square, was chosen for this study and a range of common fuels that could be found in and around a residence were selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured and flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels, in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. Thermocouple is ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer.  The response time constant is approximately 0.2~seconds~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15~\%.

@@

The natural gas flow to the burner is measured in terms of volumetric flow with a dry test meter, that is corrected for temperature and pressure. The flow rate was determined to be 4.5 cubic feet per minute as output from the meter or 2.1~liters per second.  This yields a mass flow of natural gas of approximately 2~g/s.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../Figures/Source_Fire_Flamer}\\
  \caption[A photograph of the source fire flamer and plume instrumentation apparatus]{A photograph of the source fire flamer and plume instrumentation apparatus which was centered under the 3~m by 3~m hood of the oxygen consumption calorimeter.  In the foreground, from left to right, are the video, 35~mm SLR and thermal imaging camera.}
  \label{Fire_Flamer}
\end{figure}

@@

The gasoline pan fire has was ignited and then burned until the fuel was consumed.  Therefore the heat release rate curves had more of a ``bell shape''  as opposed to the flat steady- state peak heat release rate region of the natural gas burner.  Therefore the  gasoline average peak heat release rate was based on an average over a 30 s interval bounding the peak heat release rate of each of the 18 experiments.  This method provided an average peak heat release rate of 80~kW~$\pm$~24~kW or 30~\%.  The total heat release of the gasoline over the 300~s period after ignition was 13.6~MJ~$\pm$~0.6~MJ.

@@

Examples of the photographic images are presented in Figs.~\ref{Gas_Foam_Photos} and \ref{Gasoline_Photos}.  Figure~\ref{Gas_Foam_Photos} is a photograph of the natural gas flame taken during a period of ``steady state'' burning at approximately 80~kW and a photograph of the polyurethane foam fire taken near its peak heat release rate of approximately 60~kW.  Figure~\ref{Gasoline_Photos} presents two photographs of the same gasoline fueled fire when it was burning near its peak heat release rate of 100~kW.  The two photographs, taken 0.2~s apart, represent the range of the observed flame heights from one of the gasoline experiments from 0.65~m and 1.0~m, respectively.

@@

\begin{table}
  \centering
  \begin{tabular}{|l|c|c|}
     \hline
     Fuel	            & Mean (m)  & Range (m) \\ \hline \hline
     Natural Gas	    & 0.70	    & 0.50 to 0.98 \\
     Gasoline 	        & 0.84	    & 0.52 to 1.10 \\
     Polyurethane Foam  & 0.46	    & 0.30 to 0.78 \\
     \hline
   \end{tabular}
  \caption[Mean visible flame height and averaged minimums and maximums]{Mean visible flame height and averaged minimums and maximums (Range) of visible flame heights for each of the fuels during peak heat release rate.}
  \label{tab:Flame_Heights}
\end{table}

@@

Once the characteristics of the fires had been measured, the next step was to measure the thermal impact that the fires had on a vertical target surface.  In these experiments the target is a non-combustible wall. The non-combustible material was chosen to prevent any contribution due to combustion from the wall to the fire.  The fires were placed at predetermined, fixed positions away from the wall. The fires were prepared and conducted in the same manner as the they were in the fire characterization experiments. As a result the exposure time of the target wall to the fire was 300 seconds. In the remainder of the report the term burner will be considered synonymous with the pan for the liquid fuel or the block of solid fuel. Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and then touching the wall (0.0~m).

@@

\begin{table}
	\centering
\begin{tabular}{|c|c|c|c|c|}
	\hline Density & \multicolumn{2}{c|}{Thermal Conductivity} & \multicolumn{2}{c|}{Specific Heat}   \\
	\hline (kg/m$^3$) & Temperature ($^\circ$C) & (W/mK)  & Temperature ($^\circ$C)  & (kJ/kg $^\circ$C) \\ \hline
	\hline 737  & 24 & 0.13 & 93 & 0.28 \\
	\hline  & 205 & 0.12  & 205 & 0.30 \\
	\hline  & 316 & 0.11 & 316 & 0.32 \\
	\hline  & 425 & 0.12 & 425 & 0.34 \\
	\hline  & 538 & 0.12 &  &  \\
	\hline
\end{tabular}
\caption[Thermal properties for Marinite I]{Thermal properties for Marinite I, as reported by the manufacturer BNZ~\cite{Marinite:1997}}
\label{tab:Marinite_Thermal_Properties}
\end{table}

Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2~m to 1.2~m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples, with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensors and continued, at 20~cm intervals, up to a height of 2.2~m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo} The thermocouples where installed in small holes from the back of the wall.  The bead of the thermocouple was flush with the surface on the calcium silicate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo} For the center array, a second thermocouple was installed on the backside of the calcium silicate board, approximately 10~mm below the TC positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter.

@@

Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non combustible wall natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall, 0.61~m and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However for the burner position, 0.15~m from the wall, a decrease in plume temperatures occurs approximately 3~min after ignition.  It is worth noting that as the plume temperatures decrease the temperatures of the thermocouples installed in the wall face increased, indicating that after the wall heats up the flame pulls toward the wall.  Perhpas as a result of increased convection on the face of the wall.

@@

@@

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline {Burner} & \multicolumn{2}{c}{Natural Gas} & \multicolumn{2}{c}{Gasoline} & \multicolumn{2}{c|}{PU Foam} \\
		\hline  Position & Temp~($^\circ$C) & HF~(kW/m$^{2}$) & Temp~($^\circ$C) & HF~(kW/m$^{2}$) & Temp~($^\circ$C) & HF~(kW/m$^{2}$) 	\\
		\hline \hline 0.61~m & 95 & 5 & 106 & 7 & 71 & 6 \\
		\hline 0.31~m & 164 & 14 & 232 & 24 & 148 & 16	\\
		\hline 0.15~m & 257 & 33 & 356 & 45 & 266 & 39	\\
		\hline 0.00~m & 487 & 43 & 621 & 70 & 526 & 65 	\\
		\hline
	\end{tabular}
	\caption[Summary of instrumented non-combustible wall experiments]{Summary of averaged peak temperatures and averaged heat fluxes from the instrumented non-combustible wall experiments.  The estimated combined expanded uncertainty for temperature in these experiments was $\pm$15~\% and $\pm$8~\% for the heat flux.}
	\label{tab:TW_Summary}
\end{table}

Of course the temperatures of the thermocouples in the wall are a result of the energy transferred to the wall and the thermo-physical properties of the wall material. The properties of the calcium silicate board are simlar but not identical to gypsum wallboard.  The heat flux data is independent of the wall material, so it may be a better predictor of the fire damage to gypsum wallboard when exposed to the burners at different distances.    All of the fuels generated heat fluxes in excess of 26~kW/m$^2$.  However only the burner against the wall positions provided a combination of heat flux and temperature that had the potential to ignite the paper on the gypsum wallboard based on the handbook data.  Based on the heat flux data from Figures~\ref{NCTW_HF_Center_Comp} the natural gas burner should burn the paint and paper off of the face of the gypsum wallboard up to a height of between 0.6~m amd 0.8~m above the burner.  Similarly, the gasoline burner should generate a burn area between 0.8~m and 1.0~m high on a wall composed of gypsum wallboard.  Using the same analysis, the polyurethane foam burner would create a shorter burn region approximately 0.4~m to 0.6~m.  The experiments conducted with the instrumented gypsum wall board will provide data for comparsion with data provided in this chapter and provide insight in to where the may gypsum wall board ignite.

@@

Since these experiments were the culmination of the instrumented series of full scale tests, the use of gypsum wallboard was intended to provide some insight into the amount of fire damage that would result. These experiments proved to be challenging in terms of time and effort.  Each panel of gypsum wallboard had to be carefully drilled for thermocouple and heat flux gauge placement. Since there was only one set of heat flux gauges, this reduced the number experiments that could be conducted on a given day.  With other project demands, the number of instrumented wallboard experiments that could be conducted under the oxygen consumption calorimeter was reduced due to time constraints. The test matrix is shown in Table~\ref{tab:IWGB_experiments}. There was also a concern that these experiments would have limited value due to potential interference from the water-cooled heat flux gauges and potential physical damage caused by drilling the holes for them.

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline Density & \multicolumn{2}{c|}{Thermal Conductivity}   & \multicolumn{2}{c|}{Specific Heat}  \\
		\hline (kg/m$^3$) & Temperature ($^{\circ}$C) & (W/mK)  & Temperature ($^{\circ}$C)  & (kJ/kg $^{\circ}$C) \\ \hline
		\hline 807  & 24 	& 0.14 	& 20	& 1.09 \\
		\hline  	& 100 	& 0.15  &  		&  \\
		\hline 790 	& 20 	& 0.18	& 20 	& 0.90 \\
		\hline  	& 100 	& 0.15 	& 100 	& 0.90 \\
		\hline  	& 200 	& 0.13 	& 200 	& 0.80 \\
		\hline  	& 500 	& 0.15 	& 500	& 0.90 \\
		\hline
		\end{tabular}
		\caption[Thermal properties for 12.7~mm gypsum wallboard]{Thermal properties for 12.7~mm gypsum wallboard}
		\label{tab:Gypsum wallboard_Thermal_Properties}
		\end{table}

@@

\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline Distance from wall  	& Natural Gas 		& Gasoline			& Polyurethane Foam \\
		\hline (m) 					& Number of Tests 	& Number of Tests  	& Number of Tests 	\\ \hline
		\hline 0.61 				& 1 				& 1 				& 2 			 	\\
		\hline 0.31					& 3	 				& 1					& 2 			 	\\
		\hline 0.15					& 1				 	& 1					& 2 			 	\\
		\hline 0.08					& 0 				& 1 				& 2 	 			\\
		\hline 0.00					& 1 				& 1 				& 3 	 			\\
		\hline
	\end{tabular}
	\caption[Matrix of instrumented gypsum wallboard experiments]{Matrix of instrumented gypsum wallboard experiments}
	\label{tab:IWGB_experiments}
\end{table}

@@

At the 0.15~m position the heat flux exhibits only a small decrease to the 0.2~m and 0.4~m above the burner heat flux sensors in the centerline array.  However, the heat flux sensor positioned at 0.6~m above the burner showed an increase.  As indicated above, this data which goes against the general trend of the data, was a result of the movement of the flame.  For this experiment, a sequence of photographs, Figure~\ref{IWGBGAS6_seq}, captured the oscillatory motion of the flame, which moved from the vertical centerline toward the wall and back again.  In the process, the fire pattern on the wall was generated.

@@

Figure~\ref{IWGB_NG_patterns} provides a comparison between the area of discoloration of the paint with the natural gas burner positioned at 0.15~m away from the wall versus the area of consumed and penetrated paint/paper layer which exposed the gypsum underneath when the burner was against the wall. The peak height of the consumed and penetrated paper, i.e. the burned region, was 0.8~above the burner.  This is consistent with the prediction based on the lower heat flux ignition value found in the literature and the average peak heat flux from the non-combustible wall experiments.

@@

\begin{figure}[p]
	\centering
	\includegraphics[trim=18.0in 4.0in 21.0in 14.0in, clip=true, width=0.4\columnwidth]{../Figures/IWGB_Gas_0_15_pattern}
	\includegraphics[trim=18.0in 2.6in 18.0in 11.7in, clip=true, width=0.4\columnwidth]{../Figures/IWGB_Gas7_0_8m} \\
	\caption[Photographs of the fire effects on the instrumented gypsum wallboard exposed to the gasoline burner fires]{Photographs of the fire effects on the instrumented gypsum wallboard exposed to the gasoline burner fires. For the image on the left, the burner was positioned 0.15~m from the wall.  For the image on the right, the burner was positioned 0.8~m from the wall.}
	\label{IWGB_Gas_patterns}
\end{figure}

@@

With this set of experiments, fire patterns as the result of thermal damage were finally observed in a few cases.  To recap the crude predictions from the previous chapter, the natural gas burner had the potential to cause a fire pattern with a height between 0.6~m and 0.8~m.  With the natural gas burner against the wall, a pattern with a maximum burn area of 0.8~m in height was generated.  From Chapter 3, the gasoline had the potential to create a fire pattern with a maximum height between 0.8~m and 1.0~m.  In the single instrumented gypsum wallboard experiment with the gasoline burner located against the wall, the maximum height of the burned region was just short of 0.8~m.  The prediction for the potential height of the fire pattern that could be created with the polyurethane burner against the wall was between 0.4~m and 0.6~m.  With the polyurethane burner, three replicate experiments were conducted.  The range of the maximum heights of the burned region was 0.2~m to 0.35~m.  The average peak burned region height was 0.3~m. The uncertainty of the repeatibility of the burn patterns has not been assessed yet, but it appears that the crude predictions from Ch 3 provided a ``ballpark'' estimate of the maximum height of the burned region.

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/Cone_Cal} \\
	\caption[Sketch of the cone calorimeter]{Sketch of the cone calorimeter}
	\label{Cone_Cal}
\end{figure}

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline Incident Heat Flux & Ignition & Time to Ignition &  Fire Effect  \\
		\hline (kW/m$^2$) &  & (s)  &   \\ \hline
		\hline 25	& No 	& NA 		& Paint \& paper charred \& cracked \\
		\hline 35 	& Yes 	& 73	 	& Paint \& paper consumed \\
		\hline 50	& Yes 	& 35 	 	& Paint \& paper consumed   \\
		\hline 70	& Yes 	& 21 	 	& Paint \& paper consumed \\
		\hline 75	& Yes 	& 17 		& Paint \& paper consumed \\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard time to ignition results]{Painted gypsum wallboard time to ignition results from the cone calorimeter}
	\label{tab:Gypsum_wallboard_timetoign}
\end{table}

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|}
		\hline Incident Heat Flux & Heat Release Rate per Unit Area    \\
		\hline (kW/m$^2$) & (kW/m$^2$)  \\ \hline
		\hline 35 	& 163 	\\
		\hline 50	& 179   \\
		\hline 70	& 255  	\\
		\hline 75	& 231 	\\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard peak heat release rate per unit area results]{Painted gypsum wallboard peak heat release rate per unit area results from the cone calorimeter. Each value is based on an average of three tests.}
	\label{tab:Gypsum wallboard_HRRA}
\end{table}

>>>>

Based on the evidence observed so far, you have now developed your first hypothesis about the cause of the fire.  A radiant style, electric heater was turned on.  It was too close to the furniture and ignited the fabric on the sofa or chair.  You collect the remains of the heater for further examination to have a lab determine if the heater switch was in the ``on'' position.  Now you need to start testing your hypothesis.  Is everything consistent with the other fire damage?

@@

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover, or early growth stage, fires.  A fixed fire source footprint, 30.5~cm square, was chosen for this study and a range of common fuels that could be found in and around a residence were selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured.  In addition, flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels, in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. The thermocouple was ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer.  The response time constant is approximately 0.2~seconds~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15~\%.

@@

The natural gas flow to the burner was measured in terms of volumetric flow with a dry test meter that was corrected for temperature and pressure. The flow rate was determined to be 2.1~liters per second.  This yielded a mass flow of natural gas of approximately 2~g/s.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../Figures/Source_Fire_Flamer}\\
  \caption[A photograph of the source fire flame and plume instrumentation apparatus]{A photograph of the source fire flamer and plume instrumentation apparatus which was centered under the 3~m by 3~m hood of the oxygen consumption calorimeter.  In the foreground, from left to right, are the video, 35~mm SLR and thermal imaging camera.}
  \label{Fire_Flamer}
\end{figure}

@@

The gasoline pan fire was ignited and then burned until the fuel was consumed.  Therefore the heat release rate curves had more of a ``bell shape''  as opposed to the flat steady- state peak heat release rate region of the natural gas burner.  Therefore the  gasoline average peak heat release rate was based on an average over a 30 s interval bounding the peak heat release rate of each of the 18 experiments.  This method provided an average peak heat release rate of 80~kW~$\pm$~24~kW or 30~\%.  The total heat release of the gasoline over the 300~s period after ignition was 13.6~MJ~$\pm$~0.6~MJ.

@@

Examples of the photographic images are presented in Figs.~\ref{Gas_Foam_Photos} and \ref{Gasoline_Photos}.  Figure~\ref{Gas_Foam_Photos} has photographs of the natural gas flame taken during a period of ``steady state'' burning at approximately 80~kW, the gasoline taken near its average peak heat release rate of 80~kW, and the polyurethane foam fire taken near its peak heat release rate of approximately 60~kW.  Figure~\ref{Gasoline_Photos} presents two photographs of the same gasoline fueled fire when it was burning near its peak heat release rate of 100~kW.  The two photographs, taken 0.2~s apart, represent the range of the observed flame heights from one of the gasoline experiments from 0.65~m and 1.0~m, respectively.

@@

\begin{table}
  \centering
  \begin{tabular}{|l|c|c|}
     \hline
     Fuel	            & Mean (m)  & Range (m) \\ \hline 
     Natural Gas	    & 0.70	    & 0.50 to 0.98 \\
     Gasoline 	        & 0.84	    & 0.52 to 1.10 \\
     Polyurethane Foam  & 0.46	    & 0.30 to 0.78 \\
     \hline
   \end{tabular}
  \caption[Mean visible flame height and averaged minimums and maximums]{Mean visible flame height and averaged minimums and maximums (Range) of visible flame heights for each of the fuels during peak heat release rate.}
  \label{tab:Flame_Heights}
\end{table}

@@

Once the characteristics of the fires had been measured, the next step was to measure the thermal impact that the fires had on a vertical target surface.  In these experiments the target is a non-combustible wall. The non-combustible material was chosen to prevent any contribution due to combustion from the wall to the fire.  The fires were placed at predetermined, fixed positions away from the wall. The fires were prepared and conducted in the same manner as they were in the fire characterization experiments. As a result the exposure time of the target wall to the fire was 300 seconds. In the remainder of the report the term burner will be considered synonymous with the pan for the liquid fuel or the block of solid fuel. Experiments were conducted with the closest edge of the burner located in the following positions from the wall: two burner widths (0.610~m), one burner width (0.305~m), one half burner width (0.150~m), and then touching the wall (0.0~m).

@@

\begin{table}[h]
	\centering
\begin{tabular}{|c|c|c|c|c|}
	\hline Density & \multicolumn{2}{c|}{Thermal Conductivity} & \multicolumn{2}{c|}{Specific Heat}   \\
	\hline (kg/m$^3$) & Temperature ($^\circ$C) & (W/mK)  & Temperature ($^\circ$C)  & (kJ/kg $^\circ$C) \\ \hline
	\hline 737  & 24 & 0.13 & 93 & 0.28 \\
	\hline  & 205 & 0.12  & 205 & 0.30 \\
	\hline  & 316 & 0.11 & 316 & 0.32 \\
	\hline  & 425 & 0.12 & 425 & 0.34 \\
	\hline  & 538 & 0.12 &  &  \\
	\hline
\end{tabular}
\caption[Thermal properties for Marinite I]{Thermal properties for Marinite I, as reported by the manufacturer BNZ~\cite{Marinite:1997}}
\label{tab:Marinite_Thermal_Properties}
\end{table}

Two vertical arrays of water cooled, Schmidt-Boelter total heat flux gauges, as previously described in the instrumentation section above, were installed at 20 cm intervals from 0.2~m to 1.2~m above the burner.  One vertical array was centered on the burner and the other vertical array was centered on the right edge of the burner. The arrangement is shown in Fig.~\ref{Instrumented_NC_Wall}. Bare-bead, Chromel-Alumel (type K) thermocouples, with a 0.5~mm nominal diameter were installed next to the center of each heat flux sensors and continued, at 20~cm intervals, up to a height of 2.2~m above the surface of the burner. Photographs of the front and rear of the instrumented wall are shown in Figures~\ref{Instrumented_Wall_Front_photo} and \ref{Instrumented_Wall_Rear_photo} The thermocouples where installed in small holes from the back of the wall.  The bead of the thermocouple was flush with the surface on the calcium silicate board and was in contact with the board, as shown in Figure~\ref{Instrumented_Wall_Close_up_TC_HF_photo} For the center array, a second thermocouple was installed on the backside of the calcium silicate board, approximately 10~mm below the thermocouple positioned on the front surface of the wall. The photograph in Figure~\ref{Instrumented_Wall_Detail_Rear_TC_photo} shows a rear thermocouple with the bead in contact with the rear surface of the wall and the relative position to the thermocouple on the front surface of the wall. In a manner similar to the source fire characterization experiments, the instrumented wall experiments were conducted under the NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter.

@@

\subsection{Natural Gas Burner Plume and Wall Exposures}
Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non combustible wall natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall, 0.61~m and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However for the burner position, 0.15~m from the wall, a decrease in plume temperatures occurs approximately 3~min after ignition.  It is worth noting that as the plume temperatures decrease the temperatures of the thermocouples installed in the wall face increased, indicating that after the wall heats up the flame pulls toward the wall.  Perhpas as a result of increased convection on the face of the wall.

@@

\subsection{Wall Exposures from Source Fires}

@@

\section{Fire Pattern Prediction}

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline {Burner} & \multicolumn{2}{c}{Natural Gas} & \multicolumn{2}{c}{Gasoline} & \multicolumn{2}{c|}{PU Foam} \\
		\hline  Position & Temp~($^\circ$C) & HF~(kW/m$^{2}$) & Temp~($^\circ$C) & HF~(kW/m$^{2}$) & Temp~($^\circ$C) & HF~(kW/m$^{2}$) 	\\
		\hline 0.61~m & 95 & 5 & 106 & 7 & 71 & 6 \\
		\hline 0.31~m & 164 & 14 & 232 & 24 & 148 & 16	\\
		\hline 0.15~m & 257 & 33 & 356 & 45 & 266 & 39	\\
		\hline 0.00~m & 487 & 43 & 621 & 70 & 526 & 65 	\\
		\hline
	\end{tabular}
	\caption[Summary of instrumented non-combustible wall experiments]{Summary of averaged peak temperatures and averaged heat fluxes from the instrumented non-combustible wall experiments.  The estimated combined expanded uncertainty for temperature in these experiments was $\pm$15~\% and $\pm$8~\% for the heat flux.}
	\label{tab:TW_Summary}
\end{table}

Of course the temperatures of the thermocouples in the wall are a result of the energy transferred to the wall and the thermo-physical properties of the wall material. The properties of the calcium silicate board are simlar but not identical to gypsum wallboard.  The heat flux data is independent of the wall material, so it may be a better predictor of the fire damage to gypsum wallboard when exposed to the burners at different distances.    All of the fuels generated heat fluxes in excess of 26~kW/m$^2$.  However only the burner against the wall positions provided a combination of heat flux and temperature that had the potential to ignite the paper on the gypsum wallboard based on the handbook data.  Based on the heat flux data from Figures~\ref{NCTW_HF_Center_Comp} the natural gas burner should burn the paint and paper off of the face of the gypsum wallboard up to a height of between 0.6~m amd 0.8~m above the burner.  Similarly, the gasoline burner should generate a burn area between 0.8~m and 1.0~m high on a wall composed of gypsum wallboard.  Using the same analysis, the polyurethane foam burner would create a shorter burn region approximately 0.4~m to 0.6~m.  The experiments conducted with the instrumented gypsum wall board will provide data for comparsion with data provided in this chapter and provide insight in to where the gypsum wall board may ignite.

@@

Since these experiments were the culmination of the instrumented series of full scale tests, the use of gypsum wallboard was intended to provide some insight into the amount of fire damage that would result. These experiments proved to be challenging in terms of time and effort.  Each panel of gypsum wallboard had to be carefully drilled for thermocouple and heat flux gauge placement. Since there was only one set of heat flux gauges, this reduced the number experiments that could be conducted on a given day.  With other project demands, the number of instrumented wallboard experiments that could be conducted under the oxygen consumption calorimeter was reduced due to time constraints. There was also a concern that these experiments would have limited value due to potential interference from the water-cooled heat flux gauges and potential physical damage caused by drilling the holes for them.

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|}
		\hline Density & \multicolumn{2}{c|}{Thermal Conductivity}   & \multicolumn{2}{c|}{Specific Heat}  \\
		\hline (kg/m$^3$) & Temperature ($^{\circ}$C) & (W/mK)  & Temperature ($^{\circ}$C)  & (kJ/kg $^{\circ}$C) \\ 
		\hline 807  & 24 	& 0.14 	& 20	& 1.09 \\
		\hline  	& 100 	& 0.15  &  		&  \\
		\hline 790 	& 20 	& 0.18	& 20 	& 0.90 \\
		\hline  	& 100 	& 0.15 	& 100 	& 0.90 \\
		\hline  	& 200 	& 0.13 	& 200 	& 0.80 \\
		\hline  	& 500 	& 0.15 	& 500	& 0.90 \\
		\hline
		\end{tabular}
		\caption[Thermal properties for 12.7~mm gypsum wallboard]{Thermal properties for 12.7~mm gypsum wallboard}
		\label{tab:Gypsum wallboard_Thermal_Properties}
		\end{table}

@@

\begin{table}[ht!]
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline Distance from wall  	& Natural Gas 		& Gasoline			& Polyurethane Foam \\
		\hline (m) 					& Number of Tests 	& Number of Tests  	& Number of Tests 	\\ 
		\hline 0.61 				& 1 				& 1 				& 2 			 	\\
		\hline 0.31					& 3	 				& 1					& 2 			 	\\
		\hline 0.15					& 1				 	& 1					& 2 			 	\\
		\hline 0.08					& 0 				& 1 				& 2 	 			\\
		\hline 0.00					& 1 				& 1 				& 3 	 			\\
		\hline
	\end{tabular}
	\caption[Matrix of instrumented gypsum wallboard experiments]{Matrix of instrumented gypsum wallboard experiments}
	\label{tab:IWGB_experiments}
\end{table}

@@

At the 0.15~m position the heat flux exhibits only a small decrease to the heat flux sensors 0.2~m and 0.4~m above the burner in the centerline array.  However, the heat flux sensor positioned at 0.6~m above the burner showed an increase.  As indicated above, this data which goes against the general trend of the data, was a result of the movement of the flame.  For this experiment, a sequence of photographs, Figure~\ref{IWGBGAS6_seq}, captured the oscillatory motion of the flame, which moved from the vertical centerline toward the wall and back again.  In the process, the fire pattern on the wall was generated.

@@

Figure~\ref{IWGB_NG_patterns} provides a comparison between the area of discoloration of the paint with the natural gas burner positioned at 0.15~m away from the wall versus the area of consumed and penetrated paint/paper layer which exposed the gypsum underneath when the burner was against the wall. The peak height of the consumed and penetrated paper, i.e. the burned region, was 0.8~above the burner.  This is consistent with the Chapter 3 prediction based on the lower heat flux ignition value found in the literature and the average peak heat flux from the non-combustible wall experiments.

@@

\begin{figure}[p]
	\centering
	\includegraphics[trim=18.0in 4.0in 21.0in 14.0in, clip=true, width=0.4\columnwidth]{../Figures/IWGB_Gas_0_15_pattern}
	\includegraphics[trim=18.0in 2.6in 18.0in 11.7in, clip=true, width=0.4\columnwidth]{../Figures/IWGB_Gas7_0_8m} \\
	\caption[Photographs of the fire effects on the instrumented gypsum wallboard exposed to the gasoline burner fires]{Photographs of the fire effects on the instrumented gypsum wallboard exposed to the gasoline burner fires. For the image on the left, the burner was positioned 0.15~m from the wall.  For the image on the right, the burner was positioned 0.0~m from the wall.}
	\label{IWGB_Gas_patterns}
\end{figure}

@@

With this set of experiments, fire patterns as the result of thermal damage were finally observed in a few cases.  To recap the crude predictions from the previous chapter, the natural gas burner had the potential to cause a fire pattern with a height between 0.6~m and 0.8~m.  With the natural gas burner against the wall, a pattern with a maximum burn area of 0.8~m in height was generated.  From Chapter 3, the gasoline had the potential to create a fire pattern with a maximum height between 0.8~m and 1.0~m.  In the single instrumented gypsum wallboard experiment with the gasoline burner located against the wall, the maximum height of the burned region was just short of 0.8~m.  The prediction for the potential height of the fire pattern that could be created with the polyurethane burner against the wall was between 0.4~m and 0.6~m.  With the polyurethane burner, three replicate experiments were conducted.  The range of the maximum heights of the burned region was 0.2~m to 0.35~m.  The average peak burned region height was 0.3~m. The uncertainty of the repeatibility of the burn patterns has not been assessed yet, but it appears that the crude predictions from Chapter 3 provided a ``ballpark'' estimate of the maximum height of the burned region.

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/Cone_Cal} \\
	\caption[Sketch of the cone calorimeter]{Sketch of the cone calorimeter~\cite{babrauskas:1984}}
	\label{Cone_Cal}
\end{figure}

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline Incident Heat Flux & Ignition & Time to Ignition &  Fire Effect  \\
		\hline (kW/m$^2$) &  & (s)  &   \\ 
		\hline 25	& No 	& NA 		& Paint \& paper charred \& cracked \\
		\hline 35 	& Yes 	& 73	 	& Paint \& paper consumed \\
		\hline 50	& Yes 	& 35 	 	& Paint \& paper consumed   \\
		\hline 70	& Yes 	& 21 	 	& Paint \& paper consumed \\
		\hline 75	& Yes 	& 17 		& Paint \& paper consumed \\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard time to ignition results]{Painted gypsum wallboard time to ignition results from the cone calorimeter}
	\label{tab:Gypsum_wallboard_timetoign}
\end{table}

@@

\begin{table}
	\centering
	\begin{tabular}{|c|c|}
		\hline Incident Heat Flux & Heat Release Rate per Unit Area    \\
		\hline (kW/m$^2$) & (kW/m$^2$)  \\ 
		\hline 35 	& 163 	\\
		\hline 50	& 179   \\
		\hline 70	& 255  	\\
		\hline 75	& 231 	\\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard peak heat release rate per unit area results]{Painted gypsum wallboard peak heat release rate per unit area results from the cone calorimeter. Each value is based on an average of three tests.}
	\label{tab:Gypsum wallboard_HRRA}
\end{table}
